\section{Complexity Analysis}\label{sec:analysis}

In this section, we assume a synchronous communication model with
point-to-point message passing. In this model, all nodes start
executing the algorithm simultaneously and time is divided into
synchronous rounds, i.e., when a message is sent from node $v$ to
its neighbor $u$ at time-slot $t$, it must arrive at $u$ before
time-slot $t+1$.

We first analyze the message and time complexity of the IPv6 address
allocation phase of Matrix. Then, we look into the message
complexity of the control plane of Matrix after the network
initialization phase.

Note that Matrix requires that an underlying acyclic topology (Ctree) has
been constructed by the network before the address allocation starts, i.e.,
every node knows who its parent in the Ctree is.
Moreover, one of the building blocks of Matrix is the address allocation phase, described in Section~\ref{subsec:mhcl}.

\begin{theorem} For any network of size $n$ with a
spanning collection tree $Ctree$ rooted at node $root$, the message and time
complexity of Matrix protocol in the address allocation phase is
Msg$(Matrix^{IP}$ $(Ctree, root))$ = $O(n)$ and Time $(Matrix^{IP}$ $(T, root))$
= $O(depth(Ctree))$, respectively. This message and time complexity is
asymptotically optimal.
\end{theorem}
\begin{proof}
The address allocation phase is comprised of a tree broadcast and a tree
convergecast.
In the broadcast operation, a message (with address allocation information) must be
sent to every node by the respective parent, which needs $\Omega(n)$ messages.
Moreover the message sent by the root must reach every node at distance
$depth(Ctree)$ hops away, which needs $\Omega(depth(Ctree))$ time-slots.
Similarly, in the convergecast operation, every node must send a message to its
parent after having received a message from its children, which needs
$\Omega(n)$ messages. Also, a message sent by every leaf node must reach the
root, at distance $\leq depth(Ctree)$, which needs $\Omega(depth(Ctree))$
time-slots.
\end{proof}

Next, we examine the communication cost of the routines involved in
the alternative routing, performed in the presence of persistent
node and link failures.

\begin{theorem} Consider a network with $n$ nodes and a failure event that
causes $\mathcal{L_{CT}}$ links to change in the collection tree Ctree for at
most $\Delta$ ms.
Moreover, consider a beacon interval of $\delta$ ms.
The control message complexity of Matrix to perform
alternative routing is $Msg(Matrix^{RC})$ = $O(n)$.
\end{theorem}
\begin{proof}
Consider the $\mathcal{L_{CT}}$ link changes in the collection tree
Ctree. Note that $\mathcal{L_{CT}} = O(n)$ since Ctree is acyclic
and, therefore, has at most $n-1$ links. Every link that was changed
must be inserted in the RCtree table of the respective (new) parent
and kept during the interval $\Delta$ using regularly sent beacons
from the child to the parent. Given a beacon interval of $\delta$,
the total number of control messages is bounded by
${\Delta}/{\delta} \times \mathcal{L_{CT}} = O(n)$.\end{proof}

Note that, in reality, the assumptions of synchrony and
point-to-point message delivery do not hold in a 6LoWPAN. The moment
in which each node joins the tree varies from node to node, such
that nodes closer to the root tend to start executing the address
allocation protocol earlier than nodes farther away from the root.
Moreover, collisions, node and link failures can cause delays and
prevent messages from being delivered. We analyze the performance of
Matrix in an asynchronous model with collisions and transient node
and link failures of variable duration through simulations in
Section~\ref{sec:results}.
