\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}

\usepackage{graphicx}
\usepackage{graphicx,url}
\usepackage{subfigure}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{comment}
\usepackage{algpseudocode,algorithm}
\usepackage{pifont}
\usepackage{xcolor}
\newcommand\tab[1][0.5cm]{\hspace*{#1}}
\usepackage[utf8]{inputenc}
\usepackage{amsthm}
\usepackage{tikz}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;} 
\newtheorem{theorem}{Theorem}


\modulolinenumbers[5]

%TODO: defnir Journal
\journal{Computer Communications}
%\journal{Computer Networks}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}

\begin{document}

\begin{frontmatter}

\title{Matrix: Multihop Address allocation and dynamic any-To-any Routing for
6LoWPAN}
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

\author[ufmgaddress]{Bruna Peres\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{bperes@dcc.ufmg.br}

\author[ufmgaddress]{Otavio A. de O. Souza}
\ead{oaugusto@dcc.ufmg.br}

\author[ufmgaddress]{Bruno P. Santos}
\ead{bruno.ps@dcc.ufmg.br}

\author[ufmgaddress]{Olga Goussevskaia}
\ead{olga@dcc.ufmg.br}

\author[ufmgaddress]{Marcos A. M. Vieira}
\ead{mmvieira@dcc.ufmg.br}

\author[ufmgaddress]{Luiz F. M. Vieira}
\ead{lfvieira@dcc.ufmg.br}

\author[ufmgaddress]{Antonio A. F. Loureiro}
\ead{loureiro@dcc.ufmg.br}

\address[ufmgaddress]{Computer Science Department, Universidade Federal de Minas Gerais (UFMG)\\ Av. Antonio Carlos 6627, Belo Horizonte, MG, Brazil.}



\begin{abstract}
Standard routing protocols for IPv6 over Low power Wireless Personal
Area Networks (6LoWPAN) are mainly designed for data collection
applications and work by establishing a tree-based network topology,
which enables packets to be sent upwards, from the leaves to the
root, adapting to dynamics of low-power communication links. 
In this work\footnote{This manuscript is based on preliminary conference versions \cite{peres:2016, mhclSBRC}.}, we propose Matrix, a platform-independent
routing protocol that utilizes the existing tree structure of the
network to enable reliable and efficient any-to-any data traffic in 6LoWPAN.
Matrix uses hierarchical IPv6 address assignment in order to
optimize routing table size, while preserving bidirectional routing.
Moreover, it uses a local broadcast mechanism to forward messages to
the right subtree when persistent node or link failures occur. We
implemented Matrix on TinyOS and evaluated its performance both
analytically and through simulations on TOSSIM. Our results showed
that the proposed protocol is superior to available protocols for 6LoWPAN, when it
comes to any-to-any data communication, in terms of reliability, message efficiency, and memory footprint.
\end{abstract}

\begin{keyword}
6LoWPAN \sep IPv6 \sep CTP \sep RPL \sep any-to-any routing \sep fault tolerance
\end{keyword}


\end{frontmatter}

\linenumbers

\input{intro}
\input{matrix}
\input{analysis}
\input{results}
\input{related}
\section*{Acknowledgements }
This work was supported in part by CAPES, CNPq and FAPEMIG.
\input{conclusions}

\bibliography{mybib}

\end{document}