\section{Related Work}
\label{sec:related}

AODV~\cite{perkins2003ad} and DSR~\cite{johnson2007rfc} are
traditional wireless protocols that allow any-to-any communication,
but they were designed for 802.11 and require too many states or
apply several overheads on the packet header.
In the context of low-power and lossy networks, CTP~\cite{Fonseca:2009} and
CodeDrip~\cite{junior2014codedrip} were designed for bottom-up and
top-down data flows, respectively. They support communication in only one
direction.

State-of-the-art routing protocols for 6lowPAN that enable
any-to-any communication are RPL~\cite{rfc6550}, XCTP~\cite{xctp}, and
Hydro~\cite{hydro}. RPL allows two modes of operation (storing and
non-storing) for downwards data flows. The non-storing mode is based
on source routing, and the storing mode pro-actively maintains an
entry in the routing table of every node on the path from the root
to each destination, which is not scalable to even moderate-size
networks. XCTP is an extension of CTP and is based on a reactive
reverse collection route creating between the root and every source
node. An entry in the reverse-route table is kept for every data
flow at each node on the path between the source and the
destination, which is also not scalable in terms of memory
footprint. Hydro protocol, like RPL, is based on a DAG
(directed acyclic graph) for bottom-up communication. Source nodes
need to periodically send reports to the border router, which builds
a global view (typically incomplete) of the network topology. Table \ref{tab:comparative} shows a comparison between the cited protocols for 6LoWPANs.

Some more recent protocols \cite{Palani2015, Moghadam:2015:MMR:2766739.2766774,
7374975} modified RPL to include new features. In~\cite{Palani2015}, a
load-balance technique is applied over nodes to decrease power consumption. In
\cite{Moghadam:2015:MMR:2766739.2766774, 7374975}, they provide multipath
routing protocols to improve throughput and fault tolerance.


\begin{table}[]
\centering
\caption{Comparison between related protocols for 6LoWPAN.}
\label{tab:comparative}
\begin{tabular}{@{}cccclc@{}}
\toprule
\textbf{}                   & \textbf{MATRIX} & \textbf{RPL}                 & \textbf{CTP} & \textbf{XCTP}        \\ \midrule
\textbf{Bottom-up traffic}          & \checkmark    & \checkmark                      & \checkmark   & \checkmark           \\
\textbf{Top-down traffic}           & \checkmark      & \checkmark                     &              & \checkmark           \\


\textbf{Any-to-any traffic}           & \checkmark      & \checkmark                     &              &             \\

\textbf{Address-allocation} & \checkmark      &                               & \multicolumn{1}{l}{}
 \\ 
 
 
 \textbf{Memory efficiency}           & \checkmark      &                      &  \checkmark            &             \\
 
 
 
 \textbf{Fault tolerance}           & \checkmark      &                      &              & \checkmark            \\
 
 
 
 \bottomrule
\end{tabular}
\end{table}