---
title: 'Growing up'
date: '2015-08-16'
#modified: `r format(Sys.time(), '%Y-%m-%d')`
excerpt: 'Using classes in R'
layout: post
published: yes
status: processed
comments: true
categories: [Rstats]
tags: [R, machine learning, linear regression, regularisation]
---

```{r,load packages,echo=FALSE,message=FALSE,warning=FALSE}

knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  echo = TRUE
  )

checkpoint::setSnapshot("2015-08-01")
#checkpoint::checkpoint("2015-08-01", use.knitr = TRUE)

library(dplyr)
library(magrittr)
#library(boot)
#library(ucminf)
library(ggplot2)
library(tidyr)
library(MASS)
#library(RColorBrewer)

```

I've been playing with implementations of linear and logistic regression over the last couple of months, following the exercises from a machine learning course that I have been doing. So far I have been writing things in a very functional way, constantly defining specific functions to do what are essentially generic things. 

I've also started to write a couple of my own packages, one of which I have published on [github](https://github.com/ivyleavedtoadflax/gapAPI/tree/v0.1.0) and [zenodo](https://zenodo.org/record/19743#.Vce8VPMViko). This one provides access to an API which allows users to download data collected by GPS collars on cattle. This package is entirely written as functions.

So, I think that it is about time that I started to mature my R programming career, and start to write in classes. I've been following the excellent [guide](https://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf) hosted on cran, by Friedrich Leisch, and in this post, I'm going to re-factor some of my previously functional code for linear regression with regularisation by implementing it with classes, and the usual model related methods like `print()`, `summary()`, and `predict()`.

## Before

So what does it look like at present? Rather than using the normal equation, this code uses the function `optim` to solve the problem:

$$
\min_{\theta}\frac{1}{2m}\Big(\sum^{m}_{i=1}(h_{\theta}(x^{(i)})-y^{(i)})^2\Big)+\frac{\lambda}{2m}\Big(\sum^{n}_{j=1}\theta^2_j\Big)
$$

`optim` takes in two functions, the cost function $J_\theta$ (`J`):

```{r}

J <- function(X, y, theta, lambda) {
  
  m <- length(y)
  
  theta1 <- theta
  
  # Ensure that regularisation is not operating on \theta_0
  
  theta1[1] <- 0
  
  error <- tcrossprod(theta,X)
  error <- as.vector(error) - y
  error1 <- crossprod(error,error)
  
  reg <- (lambda/(2*m)) * crossprod(theta1, theta1)
  
  cost <- (1/(2 * m)) * error1 + reg
  
  return(cost)
  
}

```

and a gradient function (`gR`):

```{r}

gR <- function(X, y, theta, lambda) {
  
  theta1 <- theta
  theta1[1] <- 0
  
  m <- length(y)
  
  error <- tcrossprod(theta,X)
  error <- as.vector(error) - y
  error <- (1/m) * crossprod(error,X)
  
  reg <- (lambda/(m)) * theta1
  
  delta <- error + reg
  
  return(delta)
  
}

```


These are taken as inputs in to the `optim` function:


```{r}

theta <- rep(1,2)
lambda <- 0
x <- mpg$displ
y <- mpg$hwy

optim_out <- optim(
  par = theta,
  fn = function(t) J(cbind(1, x), y, t, lambda),
  gr = function(t) gR(cbind(1, x), y, t, lambda),
  method = "BFGS"
)

```


And calling this, we get:


```{r}
optim_out
```


Just to check, I compare this with a simple linear regression with `lm()`.


```{r,echo = TRUE}
lm_model <- lm(hwy ~ displ, data = mpg)
coef(lm_model)
```


So the coefficients are not the same, but are pretty close nonetheless.


```{r}
all.equal(
  as.numeric(optim_out$par),
  as.numeric(coef(lm_model))
)
```


## Converting to S3 classes and methods

### Model function

As mentioned, I'm following this [guide]([guide](https://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf)) pretty closely, as it does exactly what I want to do. So I will start by producing a function to wrap up the `optim` call. `optim` must take inputs of functions for the cost and gradient, and so there is not too much that can be done to these functions as far as I can see, so for the moment I will leave them be.


```{r}

vlrr_est <- function(x, y, lambda)
  
{
  
  theta <- rep(1,ncol(x))
  
  optim_out <- optim(
    par = theta,
    fn = function(t) J(x, y, t, lambda),
    gr = function(t) gR(x, y, t, lambda),
    method = "BFGS"
  )
  
  df = nrow(x) - ncol(x)
  
  # Calculate the variance
  
  sigma2 <- sum((y - x %*% optim_out$par)^2) / df
  
  # Compute covariance with sigma^2 * (X^{T}X)^-1. Note that Leisch counsels 
  # against using this method of calculating the covariance matrix. Instead he 
  # recommends using QR decomposition. As my present implementation does not 
  # solve the normal equation (X^{T}X)^{−1}X^{T}y but instead relies on optim I 
  # do not have the products of the QR decomposition to use in the matrix 
  # inverse in the standard lm implementation: vcov <- sigma2 * chol2inv(qx$qr).
  # This implementation will do for now... Note that this requires MASS::ginv().
  # Mildly faster to use crossprod than %*%
  
  vcov <- sigma2 * ginv(crossprod(x,x))
  
  # Create a standardised output similar to that of lm()
  
  list(
    coefficients = optim_out$par,
    error = optim_out$value,
    convergence = optim_out$convergence,
    message = optim_out$message,
    vcov = vcov,
    sigma = sqrt(sigma2),
    df = df
    )
}

```


For now, `x` must be specified as the usual $\mathbb{R}^{m \times n+1}$ matrix consisting of a column of 1s which correspond to the intercept. 


```{r}
vlrr_est(x = cbind(1,mpg$displ), y = mpg$hwy, lambda = 0)
```


So far so good. Now define the function `vlrr()` and the default method for it:

### vlrr class and default method

```{r}

# This is almost entirely lifted from Leisch (2009)

vlrr <- function(x, y, lambda, ...) UseMethod("vlrr")

vlrr.default <- function(x, y, lambda = 0, ...) {
  
  # Could put in some more verbose checks of the input here.
  
  x <- as.matrix(x)
  y <- as.numeric(y)
  lambda <- as.numeric(lambda)
  
  est <- vlrr_est(x, y, lambda)
  
  est$lambda = lambda
  est$fitted.values <- as.vector(x %*% est$coefficients)
  est$residuals <- y - est$fitted.values 
  est$call <- match.call()

  class(est) <- "vlrr"
  est
}

```


So now if I call call `vlrr` or the `vlrr.default` function, I should get the same right?


```{r}
identical(
  vlrr.default(cbind(1, mpg$disp), mpg$hwy, 0), 
  vlrr(cbind(1, mpg$disp), mpg$hwy, 0)
)

```


And because I defined the output in a standard way, I can use the standard methods to interrogate it:


```{r}

model <- vlrr(cbind(1, mpg$disp), mpg$hwy, 0)

coef(model)
fitted.values(model) %>% head
resid(model) %>% head

```


### Print method

I won't `print(model)` as at present is will print all the slots in turn:


```{r}
attributes(model)
```


This ends up being very long, so it would be good to write a print method that returns something a bit shorter.


```{r}
print.vlrr <- function(x, ...) {
  cat("Call:\n")
  print(x$call)
  cat("\nError:\n")
  print(x$error)
  cat("\nCoefficients:\n")
  print(x$coefficients)
}

print(model)
```


Better.

### Summary method

Model functions also usually have a `summary` method which prints out a nice summary of the model (e.g. `summary.lm`). For now I'll just copy the default method given by Leisch, plus additional information about `optim`'s convergence success.

```{r}

summary.vlrr <- function(object, ...) {
  
  
  # It would be nice to be able to compare the success of models from the
  # summary output. Here I include a simple mean squared error.
  
  m <- length(object$fitted.values)
  error <- object$residuals
  MSE <- as.numeric(1/m * crossprod(error, error))
  
  se <- sqrt(diag(object$vcov)) 
  tval <- coef(object) / se
  
  # Note that here the p value is for testing the null hypothesis that the
  # coefficients are different from zero

  
  TAB <- cbind(
    Estimate = coef(object),
    StdErr = se, 
    t.value = tval, 
    p.value = 2*pt(-abs(tval), df=object$df)
  )
  res <- list(
    convergence = object$convergence,
    message = object$message,
    call = object$call, 
    coefficients = TAB,
    MSE = MSE
  )
  
  class(res) <-"summary.vlrr"
  res
}

summary(model)

```

This is a bit messy, but we can define a `print` method for `summary` using the `Coefmat` function to order things nicely in the coefficients table.

```{r}
print.summary.vlrr <- function(x, ...) {
  
  cat("Vectorised linear regression by optim:\n") 
  cat("\n") 
  cat("Convergence (see ?optim):\n") 
  print(x$convergence)
  print(x$message)
    
  cat("\n") 
  cat("Call:\n") 
  
  print(x$call)
  cat("\n")
  
  cat("Lambda:\n") 
  
  print(x$lambda)
  cat("\n")
  
  printCoefmat(
    x$coefficients, 
    P.value = TRUE, 
    has.Pvalue = TRUE
  )
  
  cat("\n")
  cat("MSE:\n")
  print(x$MSE)
}

print(summary(model))
```


Great, so this looks a lot more like the output we would expect from `lm()`, plus we have some statistics on the likelihood that our coefficients are different from zero.

### Formula method

It's a bit tiresome having to specify `x=` and `y=` when in `lm()` and other models, we are able to specify a formula. In addition, if we want to include interaction terms $x_1 \times x_2$ for instance, this would need to be done in the input matrix x. By implementing a `formula` method, this can all be automated. This also gets us off the hook with having to `cbind(1,x)`.

At present I must do:


```{r,eval=FALSE}
vlrr(x=cbind(1,x),y=y)
```


But after implementing the `formula` method...


```{r}

vlrr.formula <- function(formula, data = list(), ...) {
  
  mf <- model.frame(formula = formula, data = data) 
  x <- model.matrix(attr(mf, "terms"), data = mf) 
  y <- model.response(mf)
  
  est <- vlrr.default(x, y, ...) 
  est$call <- match.call() 
  est$formula <- formula 
  est
  
}

```


...things get much  much simpler!


```{r}
model <- vlrr(hwy ~ displ,lambda = 0, data = mpg)
print(model)
```


and how about:


```{r}
summary(model)
```


### Predict method

The last method I want to implement is `predict` as this makes it much easier to use the model once fit. Again I'm relying on Leisch here.


```{r}

predict.vlrr <- function(object, newdata = NULL, ...) {
  
  # If no new data, the just present the fitted values from the model
  
  if(is.null(newdata)) {
    y <- fitted(object)
  } 
  else{
    if(!is.null(object$formula)){
      ## model has been fitted using formula interface
      x <- model.matrix(object$formula, newdata)
    }
    else{
      # newdata is just a matrix
      x <- newdata
    }  
    y <- as.vector(x %*% coef(object))
  }
  y
}


```


## Testing it out

So finally we can take this model and apply it to some other data. Here I take the model I built on fuel consumption from the `mpg` dataset, and use it to predict displacement values (`disp`) from the `mtcars` dataset. Note that displacement is given in cubic inches in `mtcars` but litres in `mpg` so needs to be multiplied by 0.0163 to put it on the same scale.


```{r,2015-08-16-testing-it-out}

# Create base plot and fit using lm

plot(
  mpg ~ I(disp * 0.0163),
  data = mtcars,
  title = "mtcars"
)

abline(
  lm(mpg~I(disp * 0.0163), data = mtcars),
  col = "blue", 
  type = "o"
  )

# Use predict method on mpg model and predict over new data

new_y <- predict(
  model, 
  newdata = list(hwy = mtcars$disp, displ = mtcars$disp * 0.0163)
)

points(
  mtcars$disp * 0.0163,
  new_y,
  col = "red",
  type = "o"
)

```

## Writing a package

So the obvious thing to do now that this has all been done, is to wrap everything up into a package. I'm not going to go through the steps to do that here because it is a little bit long winded, and much better covered in the [Leisch](https://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf) guide, and [here](http://r-pkgs.had.co.nz/) by Hadley Wickham.

It's a pretty simple process, and actually the most onerous (and arguably the most important) thing is to properly document all the functions. So I have put in a little bit of time, and produced the package `vlrr`, which is now available from a github [repo](https://github.com/ivyleavedtoadflax/vlrr).

This can be installed direct from within R using `devtools::install_github("ivyleavedtoadflax/vlrr")`. The library can then be called and run like any other.

```{r,echo=FALSE}
#rm(list=ls())
```

```{r,2015-08-16-trees-dataset}
library(vlrr)

model <- vlrr(Volume ~ Girth, data = trees)

summary(model)

plot(Volume ~ Girth,data = trees)
abline(
  model, 
  col = "red", 
  lwd = 2
  )

# Something more complicated

model1 <- vlrr(Volume~poly(Girth,degree = 3), lambda = 0.1,data = trees)
summary(model1)

lines(
  trees$Girth,
  model1$fitted.values,
  col = "blue",
  lwd = 2,
  lty = 2
  )

```



```{r}
sessionInfo()  
```
