---
title: "The Hadley Effect"
author: matt_upson
comments: yes
date: '2016-04-07'
modified: `r format(Sys.time(), '%Y-%m-%d')`
layout: post
excerpt: "What happens when Hadley reads your blog"
published: no
status: processed
tags:
- R
- Rstats
- Hadley
- Wickham
categories: Rstats
---

```{r,include=FALSE}

library(checkpoint)
checkpoint("2016-03-26")

library(readr)
library(dplyr)
library(magrittr)
library(knitr)
library(broom)
library(ggplot2)
library(purrr)
library(lubridate)


knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  echo = FALSE,
  #include = FALSE,
  cache = FALSE,
  eval=TRUE,
  #fig.width = 10,
  #fig.height = 10,
  dev = "svg"
  )


```

```{r}

# read.csv(
#   "data/2016-04-01-blog-data.csv"
# ) %>% 
#   mutate(
#     date = mdy(date),
#     avg_session_duration = period_to_seconds(hms(avg_session_duration))
#     ) %>%
#   write_csv(
#     "data/2016-04-01-blog-data.csv"
#     )

blog_view_data <- read_csv(
  "data/2016-04-01-blog-data.csv",
  col_types = list(
    col_factor(
      levels = c("ivyleavedtoadflax.github.io","machinegurning.com")
    ),
    col_datetime(),
    col_integer(),
    col_double(),
    col_integer(),
    col_integer(),
    col_integer(),
    col_double(),
    col_double()
  )
) %>%
  tidyr::gather(
    key, value, perc_bounce_rate:users
  )


```

Last week [Matt G](https://twitter.com/mammykins_) tweeted about a quick blog post I wrote about [iteratively applying models](../iterating/) using some of Hadley's packages, to which Hadley replied:

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr"><a href="https://twitter.com/mammykins_">@mammykins_</a> <a href="https://twitter.com/m_a_upson">@m_a_upson</a> the only thing it&#39;s missing is that you should store related data frames and models as list columns of s data frame</p>&mdash; Hadley Wickham (@hadleywickham) <a href="https://twitter.com/hadleywickham/status/716284174076747777">April 2, 2016</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

Now this blog is pretty new, and we haven't made much of an effort to promote it.
In the time since I transferred from the old domain name (more on this later) pretty much no-one has been looking at it (apart from my mum -- thanks mum).

So what happens when Hadley reads your blog post and mentions it on twitter?

## Quantifying the Hadley effect

Well, taking the true **Hadley Effect** to be a single day event ocurring 2nd April (we term the tail of this spike: residual Hadley Effect): it's around a **7000% increase** in unique visits. 

```{r}

bind_rows(
  blog_view_data %>%
    dplyr::filter(
      site == "machinegurning.com",
      key == "users",
      date > "2016-02-28",
      date < "2016-04-02"
    ) %>%
    summarise_each(
      funs = funs(
        mean = round(mean(.)),
        sd = round(sd(.)),
        min, max),
      value
    ),
  blog_view_data %>%
    dplyr::filter(
      site == "machinegurning.com",
      key == "users",
      date > "2016-04-02"
    ) %>%
    summarise_each(
      funs = funs(
        mean = round(mean(.)),
        sd = round(sd(.)),
        min, max),
      value
    )
) %>% 
  set_rownames(
    c(
      "pre-Hadley-Effect",
      "post-Hadley-Effect"
    )
  ) %>%
  kable



```


```{r,2016-04-3-hadley-effect,fig.height=5}

blog_view_data %>%
  dplyr::filter(
    site == "machinegurning.com",
    key == "users",
    date > "2016-02-28"
  ) %>%
  ggplot +
  aes(
    x = date,
    y = value
  ) +
  geom_point() +
  geom_line() +
  theme(
    legend.position = "top"
  ) +
  ylab("Number of unique visits") +
  xlab("Date")


```

## A bit more data

So this isn't the whole story.
I've been blogging since the end of 2014 (although I only signed up to Google Analytics in April 2015), and used to host this all using github pages (actually we are still hosted by github pages: <https://machinegurning.github.io>, just we bought a err more sensible domain). So I've actually got a bit more data with which to evaluate the Hadley effect.

From a quick glance, it seems that not only is a mention from Hadley worth about 3 times as many reads as my most popular article (seems to be [this]("../cycling-weather/") one); but his briefest of mentions convinces people to read more of the blog for longer! -- Thanks Hadley

```{r,2016-04-3-hadley-effect-facet,fig.height=10}

facet_names <- c(
  users = "Unique visits (n)",
  pages_per_session = "Average pages visited per session (n)",
  avg_session_duration = "Average visit duration (seconds)"
  )

blog_view_data %>%
  dplyr::filter(
    key %in% c(
      "users",
      "pages_per_session",
      "avg_session_duration"
    )
  ) %>%
  ggplot +
  aes(
    x = date,
    y = value,
    col = site
  ) +
  geom_line() +
  facet_wrap(
    ~key,
    scales = "free_y",
    ncol = 1,
    labeller = labeller(key = facet_names)
  ) +
  theme(
    legend.position = "top"
  ) +
  ylab("") +
  xlab("Date")


```


```{r,eval=FALSE}
sessionInfo()
```

