---
title: "Lookup tables"
author: matt_gregory
comments: yes
date: '2016-03-13'
modified: `r format(Sys.time(), '%Y-%m-%d')`
layout: post
excerpt: "Using character matching for quick lookups"
published: yes
status: processed
tags:
- R
- lookups
- education
categories: Rstats
---

{% include _toc.html %}

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE
  )
```

Character matching provides a powerful way to make lookup tables. There are more concise functions available in packages like `dplyr` that achieve the same end but it is useful to understand how they are implemented with basic subsetting.

```{r,TRUE,warning=FALSE,message=FALSE}

library(dplyr)

```

We start off by building an example dataframe.

```{r}

set.seed(1337)  # we use rnorm

pupil_data <- data.frame(
  studentid = 1:12,
  school = c("Park view", "Grange Hill", "Sweet valley"),
  superoutputarea = c("E01011949", "E01011105", "E01011333"),
  attainment = (rnorm(n = 12)),
  stringsAsFactors = FALSE  #  we can modify the class of variables later if required
)

```

If we look at the data, we notice the variable [`superoutputarea`](https://neighbourhood.statistics.gov.uk/HTMLDocs/nessgeography/superoutputareasexplained/output-areas-explained.htm) is a nine digit code that doesn't tell a human much.
We are interested in how the area relates to the socio-economic classification of typical people who live in that area or a measure of deprivation of the area.
We must convert this into the more informative proxy which can then be used in our machine learning tools later.

```{r}
pupil_data
str(pupil_data)
```

What we need is a list which contains the necessary translation for `superoutputarea`. We define that here as `lookup`. A 7-point scale is used for deprivation with Sweet Valley High in a wealthy area and Park View in a deprived area.

```{r}

lookup <- c("E01011949" = 1, "E01011105" = 3, "E01011333" = 7)

```

To convert we simply:

```{r}

lookup[pupil_data$superoutputarea]
#  if we don't want the names in the result
unname(lookup[pupil_data$superoutputarea])

```

Thus we can use this to create a new variable called `depriv`. 

```{r}

pupil_data$depriv <- NULL
pupil_data$depriv <- unname(lookup[pupil_data$superoutputarea])
pupil_data
str(pupil_data)  #  check variables are of appropriate class

```

Great, now we can use this dataframe for machine learning.
What if we have a large dataframe, are there more concise and faster ready made functions to use?
Probably but we won't elucidate that here, we just assume `dplyr` is fast as it passess to `C++`.
Plus I like `dplyr` with its nice chaining.

```{r}

mutate(pupil_data, depriv2 = unname(lookup[pupil_data$superoutputarea]))

```

Sometimes we might have a more complicated lookup table which has multiple columns of infomation.
Suppose we take our vector of attainment grades and round them to the nearest whole number.

```{r}

pupil_data <- mutate(pupil_data, grade = round(attainment, digits = 0))

grades <- pupil_data$grade

info <- data.frame(
  grade = -3:3,
  desc = c("Awful", "Rubbish", "Poor", "OK", "Satisfactory", "Good", "Awesome"),
  pass = c(F, F, F, T, T, T, T)
)

```

We want to duplicate the `info` table so that we have a row for each values in `grade`. We can do this in two ways either using `match()` and integer subsetting, or `rownames()` and character subsetting:

```{r}

# using match
id <- match(grades, info$grade)
info[id, ]

# using rownames
rownames(info) <- info$grade
info[as.character(grades), ]

```

We have matched the `grade` of the student with its appropriate descriptor and pass / fail status using a more complicated lookup table.

## Conclusion
A named character vector can act as a simple lookup table. We could even read this in from a csv file. Lookup is simple in R.

## References
* Wickham, H. (2015). Advanced R.

```{r}

sessionInfo()

```
